### 2020-05-14 Manuscript/manuscript.tex
small edits and addition of NW Laurentia basin development paragraph

Laurentia is the craton that forms the Precambrian core of North America and was a major continent throughout the majority of the Proterozoic. The paleogeographic position of Laurentia is key to the development of reconstructions of Proterozoic paleogeography including the Paleoproterozoic Nuna and Neoproterozoic Rodinia supercontinents. There is a rich record of Precambrian paleomagnetic poles from Laurentia, as well as an extensive and well-documented geologic history of tectonism. These geologic and paleomagnetic records are increasingly better constrained geochronologically and are both key to evaluating and developing paleogeographic models. These data from Laurentia provides strong-support for mobile lid plate tectonic processes operating over the past 2.2 billion years.

@@

As for the suturing of the Wyoming province to Laurentia, many models posit that it was conjoined with Hearne and associated provinces at the time of the Trans-Hudson orogeny \citep[e.g.][]{St-Onge2009a, Pehrsson2015a} or was proximal to Hearne and Superior while still undergoing continued translation up to ca. 1.80 Ga \citep{Whitmeyer2007a}. A contrasting view has been been proposed that the Wyoming province and Medicine Hat blocks were not conjoined with the other Laurentia provinces until ca. 1.72 Ga \citep{Kilian2016b}. This interpretation is argued to be consistent with geochronological constraints on monazite and metamorphic zircon indicating active collisional orogenesis associated with the Big Sky orogen on the northern margin of the craton as late as ca. 1.75 to 1.72 Ga \citep{Condit2015a} and ca. 1.72 tectonomagmatic activity in the Black Hills region \citep{Redden1990a}. However, the evidence for earlier orogenesis ca. 1.78 to 1.75 in the Black Hills \citep{Dahl1999a,Hrncir2017a}, as well as high-grade tectonism as early as ca. 1.81 Ga in the Big Sky orogen \citep{Condit2015a}, may support the interpretation of \citet{Hrncir2017a} that ca. 1.72 Ga activity is a minor overprint on ca. 1.75 terminal suturing between Wyoming and Superior. Regardless, in both of these interpretations, Wyoming is a later addition to Laurentia with final suturing post-dating ca. 1.82 Ga amalgamation of Archean provinces with the Trans-Hudson orogen further to the northeast. Overall, the collision of these Archean microcontinents between ca. 1.9 and 1.8 Ga led to rapid amalgamation of the majority of the Laurentia craton.

@@

Crustal growth also progressed in the Paleoproterozoic through accretionary orogenesis. This accretion occurred within the Wopmay orogen through ca. 1.88 Ga arc-continent collision that led to the accretion of the Hottah terrane (the Calderian orogeny) and the subsequent emplacement of the Great Bear magmatic zone from ca. 1.88 to 1.84 Ga \citep{Hildebrand2009a}. Coeval with the Trans-Hudson orogeny was the perphiral Penokean orogeny during which both microcontinent blocks (the Marshfield terrane) and arc terranes accreted on the southeastern margin of the west Superior province ca. 1.86 to 1.82 \citep{Schulz2007a}. Firm evidence of the end of the orogeny comes from the ca. 1.78 undeformed plutons of the post-Penokean East Central Minnesota Batholith \citep{Holm2005a}.

The collisions of provinces and terranes leading up to the Trans-Hudson orogeny mark the initial phase of assembly of the supercontinent Nuna in the paleogeographic model framework of \cite{Pehrsson2015a}. The Trans-Hudson orogeny itself is taken to be the terminal collision associated with the closure of the Manikewan Ocean that had previously been a large oceanic tract separating the Superior province from the composite Slave+Rae+Hearne+North Atlantic provinces (often referred to as the Churchill domain or plate; e.g. \citealp{Skipton2016a, Weller2017a}). The \cite{Pehrsson2015a} model posits that this period of terminal collision not only resulted in the amalgamation of Laurentia, but was also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Baltica, Siberia, Congo/S\~ao Francisco, West Africa, and Amazonia \citep{Whitmeyer2007a, Pehrsson2015a}.

Following the Trans-Hudson orogeny, the locus of orogenesis migrated to the exterior of Laurentia. This change marks a shift in the predominant style of Laurentia's growth as subsequent crustal growth occurred dominantly through accretion of juvenile crust along the southern and eastern margin of the nucleus of Archean provinces (\citealp{Whitmeyer2007a}; Figs. \ref{fig:Laurentia_map} and \ref{fig:tectonic_history}). Determining the extent of these belts is complicated by poor exposure of them in the midcontinent relative to the exposure of the Archean provinces throughout the Canadian shield. Major growth of Laurentia following the amalgamation of these Archean provinces occurred associated with the arc-continent collision of the ca. 1.71 to 1.68 Ga Yavapai orogeny. Yavapai orogenesis is interpreted to have resulted from the accretion of a series of arc terranes that collided with each other and Laurentia \citep{Karlstrom2001a}. Yavapai accretion was followed by widespread emplacement of granitoid intrusions \citep{Whitmeyer2007a}. These intrusions are hypothesized to have stabilized the juvenile accreted terranes that subsequently remained part of Laurentia \citep{Whitmeyer2007a}. Subsequent accretionary orogenesis of the ca. 1.65 to 1.60 Ga Mazatzal orogeny and associated plutonism led to further crustal growth in the latest Paleoproterozoic \citep{Karlstrom1988a}. Laurentia's growth continued in the Mesoproterozoic along the southeast margin through further juvenile terrane and arc accretion. An interval of major plutonism occurred ca. 1.48 to 1.35 Ga leading to the formation of A-type granitoids throughout both Mesoproterozoic and Paleoproterozoic provinces extending from the southwest United States up to the Central Gneiss Belt of Ontario to the northeast of Georgian Bay \citep{Slagstad2009a}. This plutonism is likely due to crustal melting within a back-arc region of ca. 1.50 to 1.43 Ga accretionary orogenesis \citep{Bickford2015a}. Geologic data from northern New Mexico have been interpreted to indicate an interval of ca. 1.49 to 1.40 Ga orogenesis that has been named the Picuris orogeny \citep{Daniel2013a}. Younger magmatic activity ca. 1.37 Ga of the Southern Granite-Rhyolite Province suggests a similar tectonic setting of accretionary orogenesis at that time \citep{Bickford2015a}. While an active margin interpretation with magmatism in back-arc setting has gained traction within the literature with additional data, the tectonic setting is often described as enigmatic given earlier interpretations of an anorogenic setting (see references in \citealp{Slagstad2009a}).

@@

Accretionary orogenesis continued along the east margin of Laurentia with the arc-continent collision of the ca. 1.25 to 1.22 Ga Elzevirian orogeny \citep{McLelland2013a}. The subsequent ca. 1.19 to 1.16 Ga Shawinigan orogeny is interpreted to be due to the accretion of a terrane comprising amalgamated arc volcanics and associated metasediments and is followed by a period of tectonic quiescence on the eastern margin of Laurentia until the collision orogenesis of the Grenvillian orogeny \citep{McLelland2010a}. In the latest Mesoproterozoic (ca. 1.11 to 1.08 Ga), a major intracontinental rift co-located with a large igneous province formed in Laurentia's interior leading to extension within the Archean Superior province and adjacent Paleoproterozoic provinces to the south \citep{Cannon1992b}. This Midcontinent Rift led to the formation of a thick succession of volcanics and mafic intrusions that are well-preserved in Laurentia's interior. Midcontinent Rift development ceased as major collisional orogenesis of the Grenvillian orogeny began \citep{Swanson-Hysell2019a}. The Grenvillian orogeny was a protracted interval of continent-continent collision (ca. 1.09 to 0.98 Ga) leading to amphibolite to granulite facies metamorphism through the orogen \citep{McLelland2010a}. The orogeny is interpreted to have resulted in the development of a thick plateau associated with the Ottawan orogenic phase (ca. 1090 to 1030 Ma; \citealp{Rivers2008a}). Continued convergence during the Rigolet phase of the Grenvillian orogeny led to the development of the Grenville Front tectonic zone and ended ca. 980 Ma \citep{Hynes2010a}.

@@

This late Paleoproterozoic and Mesoproterozoic tectonic history provides significant constraints on paleogeographic reconstructions. In particular, the long-lived history of accretionary orogenesis along the southeast (present-day coordinates) of Laurentia from the initiation of the Yavapai orogeny (ca. 1.71 Ga) to the end of the Shawinigan orogeny (ca. 1.06 Ga) requires a long-lived open margin without a major conjugate continent until the time of terminal Grenvillian orogeny collision \citep{Karlstrom2001a}. This constraint is incorporated into models such as that of \citet{Zhang2012a} and \citet{Pehrsson2015a} which maintain a long-lived convergent margin throughout the Mesoproterozoic, but in some reconstructions other continental blocks are reconstructed into positions that are seemingly incompatible with this record of accretionary orogenesis (e.g. Amazonia in \citealp{Elming2009a}). The high-grade metamorphism associated with the Ottawan phase of the Grenvillian orogeny itself strongly suggests a collision between Laurentia and (an)other continent(s) ca. 1080 Ma -- the geological observation of which first led to the formulation of the hypothesis of the supercontinent Rodinia \citep{Hoffman1991a}. Evidence of large-scale continent-continent collision at the time of the Ottawan Phase of the Grenvillian orogeny is recorded in Texas, up through the Blue Ridge Appalachian inliers, through Ontario and up to the Labrador Sea. This extensive and major collisional orogenic history recorded in Laurentia remains a strong piece of evidence that a supercontinent or (proto)supercontinent formed at the 1.0 Ga Mesoproterozoic to Neoproterozoic transition. The term Grenville orogeny or Grenville belt is used rather loosely throughout much of the literature to refer to any late Mesoproterozoic orogenic belt. The timeline of orogenesis on the Laurentia margin has more nuanced constraints than this usage, and these constraints can be comparatively assessed when evaluating potential conjugate continents to Laurentia associated with the orogen (Fig. \ref{fig:tectonic_history}).

The subsequent Neoproterozoic tectonic history of Laurentia is dominantly a record of rifting (Fig. \ref{fig:tectonic_history}). Along the western margin of Laurentia, rifting occurred ca. 780 to 720 Ma leading to deposition in basins from the Death Valley region of SW Laurentia to the Mackenzie Mountains of NW Laurentia \citep{Macdonald2012a, Dehler2017a, Rooney2017a}. However, this extensional basin development predates the rifting that led to well-documented passive margin thermal subsidence closer to the ca. 539 Ma Neoproterozoic-Phanerozoic boundary \citep{Bond1984a, Levy1991a}. The emplacement of the ca. 780 Ma Gunbarrel large igneous province along this margin and the subsequent extension recorded in the basins is commonly interpreted to be associated with the break-up of Laurentia and a conjugate continent to the western margin (often interpreted to be Australia). If this interpretation is correct, it is unclear why there would be minimal thermal subsidence until the Ediacaran (post 635 Ma as in \citealp{Levy1991a} and \citealp{Witkosky2018a}). The geological evidence therefore supports prolonged active tectonism along the western margin of Laurentia (a portion of which could be strike-slip and transtensional; \citealp{Smith2015b}), but suggests that there was significant lithospheric thinning associated with rifting later than the timing of rifting typically implemented in models of Rodinia break-up. The record of Neoproterozoic basin development lead \cite{Yonkee2014a} to propose that the early ca. 780 Ma rifting was intracratonic and that while it may have led to some associated thermal subsidence that there was a second interval of rifting and thermal subsidence associated with Australia rifting away in the Ediacaran (later than in most models). Another possibility, along the lines of that proposed in \citet{Ross1991a}, is that ca. 780 Ma extensional tectonism is an inboard record of rifting and passive margin development that occurred further outboard. In this model, subsequent continent rifting that drove lithospheric thinning, perhaps associated with the departure of a microcontinent fragment rather than an already departed major conjugate continent, would be the cause of Ediacaran to Cambrian thermal subsidence.

The margin that did experience large-scale rifting and associated passive margin thermal subsidence earlier in the Neoproterozoic is the northeast Greenland margin (Fig. \ref{fig:tectonic_history}). Available geochronological constraints and thermal subsidence modeling indicate ca. 820 Ma rifting followed by thermal subsidence of a stable platform \citep{Maloof2006a, Halverson2018a}. These data suggest that conjugate continental lithosphere rifted away from northeast Greenland ca. 820 Ma.

Extensive rifting followed by thermal subsidence occurred along the southeast to east Laurentia margin leading up to the Neoproterozoic-Phanerozoic boundary and is interpreted to be associated with the opening of the Iapetus ocean. A record of this rifting is preserved as rift basins that were part of failed arms (Rome trough, Reelfoot rift and Oklahoma aulacogen; Fig. \ref{fig:Laurentia_map}) as well as prolonged Cambrian to Ordovician passive margin thermal subsidence along the margin \citep{Bond1984a, Whitmeyer2007a}. The age of igneous intrusions that have been interpreted to be rift-related play a significant role in interpretations of this history such as in the rift development model of \citet{Burton2010a}. In this model, spatially-restricted rifting occurs ca. 760 to 680 Ma in the region of modern-day North Carolina and Virginia. Rifting ca. 620 to 580 Ma initiates in the region from modern-day New York to Newfoundland and by ca. 580 to 550 Ma rifting extends along the length of Laurentia's eastern margin. The last phase of this rifting appears to be associated with the separation of the Argentine pre-Cordillera Cuyania terrane \citep{Dickerson1998a}.

@@

The eastern margin of Laurentia then went through the cycle of Appalachian orogenesis. As is visualized in Figure \ref{fig:tectonic_history}, there are parallels between the Grenville orogenic interval and the Appalachian orogenic interval in that there was a period of arc-continent collision (Shawinigan orogeny in the Grenville interval; Taconic orogeny in the Appalachian interval) followed by microcontinent accretion (Llano in the Grenville interval; Acadian in the Appalachian interval) that culminated in large-scale continent-continent collision (Grenvillian orogeny in the Grenville interval; Alleghanian in the Appalachian interval). These similarities are the consequence of an active margin facing an ocean basin that was progressively consumed until continent-continent collision. In the case of the Grenville interval, this terminal collision is interpreted to be associated with the assembly of the supercontinent Rodinia, and in the Appalachian interval it is interpreted to be associated with the assembly of the supercontinent Pangea.

>>>>

Laurentia is the craton that forms the Precambrian core of North America and was a major continent throughout the majority of the Proterozoic. The paleogeographic position of Laurentia is key to the development of reconstructions of Proterozoic paleogeography including the Paleoproterozoic-Mesoproterozoic Nuna and Neoproterozoic Rodinia supercontinents. There is a rich record of Precambrian paleomagnetic poles from Laurentia, as well as an extensive and well-documented geologic history of tectonism. These geologic and paleomagnetic records are increasingly better constrained geochronologically and are both key to evaluating and developing paleogeographic models. These data from Laurentia provides strong-support for mobile lid plate tectonic processes operating continuously over the past 2.2 billion years.

@@

As for the suturing of the Wyoming province to Laurentia, many models posit that it was conjoined with Hearne and associated provinces at the time of the Trans-Hudson orogeny \citep[e.g.][]{St-Onge2009a, Pehrsson2015a} or was proximal to Hearne and Superior while still undergoing continued translation up to ca. 1.80 Ga \citep{Whitmeyer2007a}. A contrasting view has been been proposed that the Wyoming province and Medicine Hat blocks were not conjoined with the other Laurentia provinces until ca. 1.72 Ga \citep{Kilian2016b}. This interpretation is argued to be consistent with geochronological constraints on monazite and metamorphic zircon indicating active collisional orogenesis associated with the Big Sky orogen on the northern margin of the craton as late as ca. 1.75 to 1.72 Ga \citep{Condit2015a} and ca. 1.72 tectonomagmatic activity in the Black Hills region \citep{Redden1990a}. However, the evidence for earlier orogenesis ca. 1.78 to 1.75 in the Black Hills \citep{Dahl1999a,Hrncir2017a}, as well as high-grade tectonism as early as ca. 1.81 Ga in the Big Sky orogen \citep{Condit2015a}, may support the interpretation of \citet{Hrncir2017a} that ca. 1.72 Ga activity is a minor overprint on ca. 1.75 terminal suturing between Wyoming and Superior. Regardless, in both of these interpretations, Wyoming is a later addition to Laurentia with final suturing post-dating ca. 1.82 Ga amalgamation of Archean provinces with the Trans-Hudson orogen further to the northeast. Overall, the collision of these Archean microcontinents between ca. 1.9 and 1.8 Ga led to rapid amalgamation of the majority of the Laurentia craton (Fig. \ref{fig:Laurentia_map}).

@@

Crustal growth also progressed in the Paleoproterozoic through accretionary orogenesis. This accretion occurred within the Wopmay orogen through ca. 1.88 Ga arc-continent collision that led to the accretion of the Hottah terrane (the Calderian orogeny) and the subsequent emplacement of the Great Bear magmatic zone from ca. 1.88 to 1.84 Ga \citep{Hildebrand2009a}. Coeval with the Trans-Hudson orogeny was the peripheral Penokean orogeny during which both microcontinent blocks (the Marshfield terrane) and arc terranes accreted on the southeastern margin of the west Superior province ca. 1.86 to 1.82 \citep{Schulz2007a}. Firm evidence of the end of the orogeny comes from the ca. 1.78 undeformed plutons of the post-Penokean East Central Minnesota Batholith \citep{Holm2005a}.

The collisions of provinces and terranes leading up to the Trans-Hudson orogeny mark the initial phase of assembly of the supercontinent Nuna in the paleogeographic model framework of \cite{Pehrsson2015a}. The Trans-Hudson orogeny itself is taken to be the terminal collision associated with the closure of the Manikewan Ocean that had previously been a large oceanic tract separating the Superior province from the composite Slave+Rae+Hearne+North Atlantic provinces (often referred to as the Churchill domain or plate; e.g. \citealp{Skipton2016a, Weller2017a}). The paleogeographic mode of \cite{Pehrsson2015a} posits that this period of terminal collision not only resulted in the amalgamation of Laurentia, but was also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Baltica, Siberia, Congo/S\~ao Francisco, West Africa, and Amazonia.

Following the Trans-Hudson orogeny, the locus of orogenesis migrated to the exterior of Laurentia. This change marks a shift in the predominant style of Laurentia's growth as subsequent crustal growth occurred dominantly through accretion of juvenile crust along the southern and eastern margin of the nucleus of Archean provinces (\citealp{Whitmeyer2007a}; Figs. \ref{fig:Laurentia_map} and \ref{fig:tectonic_history}). Determining the extent of these belts is complicated by poor exposure of them in the midcontinent relative to the exposure of the Archean provinces throughout the Canadian shield. Major growth of Laurentia following the amalgamation of these Archean provinces occurred associated with the arc-continent collision of the ca. 1.71 to 1.68 Ga Yavapai orogeny. Yavapai orogenesis is interpreted to have resulted from the accretion of a series of arc terranes that collided with each other and Laurentia \citep{Karlstrom2001a}. Yavapai accretion was followed by widespread emplacement of granitoid intrusions \citep{Whitmeyer2007a}. These intrusions are hypothesized to have stabilized the juvenile accreted terranes that subsequently remained part of Laurentia \citep{Whitmeyer2007a}. Subsequent accretionary orogenesis of the ca. 1.65 to 1.60 Ga Mazatzal orogeny and associated plutonism led to further crustal growth in the latest Paleoproterozoic \citep{Karlstrom1988a}. Laurentia's growth continued in the Mesoproterozoic along the southeast margin through further juvenile terrane and arc accretion. An interval of major plutonism occurred ca. 1.48 to 1.35 Ga leading to the formation of A-type granitoids throughout both Mesoproterozoic and Paleoproterozoic provinces extending from the southwest United States up to the Central Gneiss Belt of Ontario to the northeast of Georgian Bay \citep{Slagstad2009a}. This plutonism is likely due to crustal melting within a back-arc region of ca. 1.50 to 1.43 Ga accretionary orogenesis \citep{Bickford2015a}. Geologic data from northern New Mexico have been interpreted to indicate an interval of ca. 1.49 to 1.40 Ga orogenesis that has been named the Picuris orogeny \citep{Daniel2013a}. Younger magmatic activity ca. 1.37 Ga of the Southern Granite-Rhyolite Province suggests a similar tectonic setting of accretionary orogenesis at that time \citep{Bickford2015a}. While an active margin interpretation with magmatism in a back-arc setting has gained traction within the literature, the tectonic setting is often described as enigmatic given earlier interpretations of an anorogenic setting (see references in \citealp{Slagstad2009a}).

@@

Accretionary orogenesis continued along the east margin of Laurentia with the amalgamation and accretion of arcs associated with the ca. 1.25 to 1.22 Ga Elzevirian orogeny \citep{McLelland2013a}. The subsequent ca. 1.19 to 1.16 Ga Shawinigan orogeny is interpreted to be due to the collision of terrane comprised of a previously rifted fragment of Laurentia and led to obduction of the Pyrites Complex ophiolite \citep{McLelland2010a, Chiarenzelli2011a}. The Shawinigan orogeny is followed by a period of tectonic quiescence on the eastern margin of Laurentia until the collisional orogenesis of the Grenvillian orogeny \citep{McLelland2010a}. In the latest Mesoproterozoic (ca. 1.11 to 1.08 Ga), a major intracontinental rift co-located with a large igneous province formed in Laurentia's interior leading to extension within the Archean Superior province and adjacent Paleoproterozoic provinces to the south \citep{Cannon1992b}. This Midcontinent Rift led to the formation of a thick succession of volcanics and mafic intrusions that are well-preserved in Laurentia's interior. Midcontinent Rift development ceased as major collisional orogenesis of the Grenvillian orogeny began \citep{Swanson-Hysell2019a}. The Grenvillian orogeny was a protracted interval of continent-continent collision (ca. 1.09 to 0.98 Ga) leading to amphibolite to granulite facies metamorphism through the orogen \citep{McLelland2010a}. The orogeny is interpreted to have resulted in the development of a thick plateau associated with the Ottawan orogenic phase (ca. 1090 to 1030 Ma; \citealp{Rivers2008a}). Continued convergence during the Rigolet phase of the Grenvillian orogeny led to the development of the Grenville Front tectonic zone and ended ca. 980 Ma \citep{Hynes2010a}.

@@

This late Paleoproterozoic and Mesoproterozoic tectonic history provides significant constraints on paleogeographic reconstructions. In particular, the long-lived history of accretionary orogenesis along the southeast (present-day coordinates) of Laurentia from the initiation of the Yavapai orogeny (ca. 1.71 Ga) to the end of the Shawinigan orogeny (ca. 1.16 Ga) requires a long-lived open margin without a major conjugate continent until the time of terminal Grenvillian orogeny collision \citep{Karlstrom2001a}. This constraint is incorporated into models such as that of \citet{Zhang2012a} and \citet{Pehrsson2015a} which maintain a long-lived convergent margin throughout the Mesoproterozoic, but in some reconstructions other continental blocks are reconstructed into positions that are seemingly incompatible with this record of accretionary orogenesis (e.g. Amazonia in \citealp{Elming2009a}). The high-grade metamorphism associated with the Ottawan phase of the Grenvillian orogeny itself strongly suggests a collision between Laurentia and (an)other continent(s) ca. 1080 Ma -- the geological observation of which first led to the formulation of the hypothesis of the supercontinent Rodinia \citep{Hoffman1991a}. Evidence of large-scale continent-continent collision at the time of the Ottawan Phase of the Grenvillian orogeny is recorded in Texas, up through the Blue Ridge Appalachian inliers \citep{Johnson2020a}, through Ontario and up to the Labrador Sea \citep{Rivers2008a}. This extensive and major collisional orogenic history recorded in Laurentia remains a strong piece of evidence that a supercontinent or (proto)supercontinent formed at the 1.0 Ga Mesoproterozoic to Neoproterozoic transition. The term Grenville orogeny or Grenville belt is used rather loosely throughout much of the literature to refer to any late Mesoproterozoic orogenic belt. The timeline of orogenesis on the Laurentia margin has more nuanced constraints than this usage, and these constraints can be comparatively assessed when evaluating potential conjugate continents to Laurentia associated with the orogen (Fig. \ref{fig:tectonic_history}).

The subsequent Neoproterozoic tectonic history of Laurentia is dominantly a record of rifting (Fig. \ref{fig:tectonic_history}). Along the western margin of Laurentia, rifting occurred ca. 780 to 720 Ma leading to deposition in basins from the Death Valley region of SW Laurentia to the Mackenzie Mountains of NW Laurentia \citep{Macdonald2013a, Dehler2017a, Rooney2017a}. However, this extensional basin development predates the rifting that led to well-documented passive margin thermal subsidence closer to the ca. 539 Ma Neoproterozoic-Phanerozoic boundary \citep{Bond1984a, Levy1991a}. The emplacement of the ca. 780 Ma Gunbarrel large igneous province along this margin and the subsequent extension recorded in the basins is commonly interpreted to be associated with the break-up of Laurentia and a conjugate continent to the western margin (often interpreted to be Australia). If this interpretation is correct, it is unclear why there would be minimal thermal subsidence until the Ediacaran (post 635 Ma as in \citealp{Levy1991a} and \citealp{Witkosky2018a}). The geological evidence therefore supports prolonged active tectonism along the western margin of Laurentia (a portion of which could be strike-slip and transtensional; \citealp{Smith2015b}), but suggests that there was significant lithospheric thinning associated with rifting later than the timing of rifting typically implemented in models of Rodinia break-up. The record of Neoproterozoic basin development lead \cite{Yonkee2014a} to propose that the early ca. 780 Ma rifting was intracratonic and that while it may have led to some associated thermal subsidence that there was a second interval of rifting and thermal subsidence associated with Australia rifting away in the Ediacaran (later than in most models). Another possibility, along the lines of that proposed in \citet{Ross1991a}, is that ca. 780 Ma extensional tectonism is an inboard record of rifting and passive margin development that occurred further outboard. In this model, subsequent continent rifting that drove lithospheric thinning, perhaps associated with the departure of a microcontinent fragment rather than an already departed major conjugate continent, would be the cause of Ediacaran to Cambrian thermal subsidence.

In northwest Laurentia, the sedimentary rock record is distinct from further south as it also records earlier Neoproterozoic basin development during the Tonian Period in addition to Cryogenian basin development \citep{Macdonald2012a}. Tectonic extension is recorded in units of the lower Fifteenmile Group with maximum depositional ages of ca. 1050 Ma with ongoing basin development ca. 812 Ma (age constraint from a U-Pb zircon date on a tuff within the upper Fifteenmile Group; \citealp{Macdonald2010a}) potentially through thermal subsidence \citep{Macdonald2012a}. Earlier basin development recorded by the Mesoproterozoic/Neoproterozoic Pinguicula Group could provide valuable insight on tectonic history as it has been interpreted to have been deposited in an extensional basin \citep{Medig2016a}, however it is poorly constrained in terms of age -- older than the Fifteenmile Group and younger than the ca. 1382 Ma Hart River sills (which themselves have been interpreted to be emplaced in conjunction with rifting; \citealp{Verbaas2018a}).

Another margin that did experienced rifting and associated passive margin thermal subsidence earlier in the Neoproterozoic is the northeast Greenland margin (Fig. \ref{fig:tectonic_history}). Available geochronological constraints and thermal subsidence modeling indicate ca. 850 to 820 Ma rifting followed by thermal subsidence of a stable platform \citep{Maloof2006a, Halverson2018a}. These data suggest that conjugate continental lithosphere had rifted away from northeast Greenland by ca. 820 Ma.

Extensive rifting followed by thermal subsidence occurred along the southeast to east Laurentia margin leading up to the Neoproterozoic-Phanerozoic boundary and is interpreted to be associated with the opening of the Iapetus ocean. A record of this rifting is preserved as rift basins that were part of failed arms (Rome trough, Reelfoot rift and Oklahoma aulacogen; Fig. \ref{fig:Laurentia_map}) as well as prolonged Cambrian to Ordovician passive margin thermal subsidence along the margin \citep{Bond1984a, Whitmeyer2007a}. The age of igneous intrusions that have been interpreted to be rift-related play a significant role in interpretations of this history such as in the rift development model of \citet{Burton2010a}. In this model, spatially-restricted rifting occurs ca. 760 to 680 Ma in the region of modern-day North Carolina and Virginia. Rifting ca. 620 to 580 Ma initiates in the region from modern-day New York to Newfoundland and by ca. 580 to 550 Ma rifting extends along the length of Laurentia's eastern margin. The last phase of this rifting has been interpreted to be associated with the separation of the Argentine pre-Cordillera Cuyania terrane \citep{Dickerson1998a}.

@@

The eastern margin of Laurentia then went through the multiple phases of Appalachian orogenesis. As is visualized in Figure \ref{fig:tectonic_history}, there are parallels between the Grenville orogenic interval and the Appalachian orogenic interval in that there was a period of arc-continent collision (Elzevirian orogeny in the Grenville interval; Taconic orogeny in the Appalachian interval) followed by microcontinent accretion (Shawinigan/Llano in the Grenville interval; Acadian in the Appalachian interval) that culminated in large-scale continent-continent collision (Grenvillian orogeny in the Grenville interval; Alleghanian in the Appalachian interval). These similarities are the consequence of an active margin facing an ocean basin that was progressively consumed until continent-continent collision. In the case of the Grenville interval, this terminal collision is interpreted to be associated with the assembly of the supercontinent Rodinia, and in the Appalachian interval it is interpreted to be associated with the assembly of the supercontinent Pangea.
